---
title: "Targeted Fst age interaction"
author: "Lea Huber"
date: '`r Sys.Date()`'
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,fig.height = 4, fig.width = 6, fig.align = 'center',fig.pos = "h")
knitr::opts_knit$set(root.dir = "/Users/Lea/Library/CloudStorage/OneDrive-Aarhusuniversitet/LeaH/Subspecies/")
pacman::p_load(tidyverse,knitr,kableExtra,ggpubr,FSA,ggbreak)
```

```{r load data}
# fst and other pop stats
setwd("/Users/Lea/Library/CloudStorage/OneDrive-Aarhusuniversitet/LeaH/Subspecies/")
Popstats <- read.csv("data/FstTargeted/monobiallelic.500bp_start.ageint.regionsOI_fst.csv")
regInfo <- read.table("data/FstTargeted/regionsOI_ageint_500bp.with_names.bed", col.names = c("scaffold","start","end","geneID", "start_codon"))
snp_counts <- read.table("data/FstTargeted/monobiallelic.500bp_start.ageint.snp_counts.bed", col.names = c("scaffold","start","end","snps"))

# add gene ids and snp counts to the pop stats df
Popstats <- full_join(Popstats,regInfo, by = join_by(scaffold,start,end))
Popstats <- full_join(Popstats,snp_counts, by = join_by(scaffold,start,end))

tempcols2 = c(shared = "#6800a4", cold = "#377EB8", hot = "#E41A1C")

# gene expression results from glmmSeq for age interaction genes
load("../ChamberTempExp/data/working/RNAseq_countdata/age_dir_cats.RData")

# add gene category to Popstats
Popstats <- full_join(Popstats, age_dirs, by = join_by(geneID == gene))

```

```{r end on scaffold, include=FALSE}
# I made sure that they don't start before position 1, but the regions could go over end of scaffold
# does that happen? What are the consequences (error or just calculates with sites that are there?)
scaff_ends <- read.table("QC/VCFstats/reference_scaffoldlength.txt", col.names = c("scaffold","scaff_end"))
Popstats <- left_join(Popstats,scaff_ends) # just add it to popstats 
which(Popstats$end > Popstats$scaff_end) # 4 of those have no data because they are on <1000 scaffolds
# one has no data for some other reason
# and two do have data, with a mid and nr of sites that makes sense, so it just takes the actual end of the region for calculation no problem
```
 
```{r filter}
# remove NA's and age interaction genes
summary(Popstats$snps)
Popstats <- filter(Popstats, !is.na(Fst_blue_red) &
                     snps > 1) # set minimum nr of snps
# how many each category?
table(Popstats$cvb)
table(Popstats$hvb)
```

Removing regions with just one SNP removes 5 rows with NAs and 5 more with just one snp.

Get some overview of the data and do some checks.  

```{r checks, include=FALSE}
# check if the start codon makes sense
plot(Popstats$end, Popstats$start_codon)
abline(0,1) # yes it always seems to be very close
plot(Popstats$end - Popstats$start_codon, ylim = c(0,3000)) # many are at 500
plot(Popstats$start_codon - Popstats$start, ylim = c(0,3000))
# looks good
```

Maybe the number of sites or the number of SNPs influences the $F_{ST}$, so let's check that. 

```{r site or snp influence, fig.show='hold', fig.width=3, fig.height=4}
plot(Popstats$snps, Popstats$Fst_blue_red)
plot(Popstats$snps, Popstats$dxy_blue_red)
```

Regions with few SNPs have a lot more variation in $F_{ST}$, maybe because then there would be more error associated with it.  

What about in interaction with SNP density, so regions with many SNP's compared to the number of sites, or few.  

```{r missing sites and snp density, fig.show='hold', fig.width=3, fig.height=4}
Popstats$missingSites <- (Popstats$end - Popstats$start) - Popstats$sites
hist(Popstats$missingSites) # there're very few negative
# which(Popstats$missingSites < 0) # 2
ggplot(Popstats, aes(x = end - start, y = sites, colour = Fst_blue_red))+
  geom_point(alpha = 0.5)+
  theme_pubr()

ggplot(Popstats, aes(x = sites, y = snps, colour = missingSites))+
  geom_point(alpha = 0.5)+
  theme_pubr()
ggplot(Popstats, aes(x = snps/sites, y = Fst_blue_red))+
  geom_point(alpha = 0.5)+
  theme_pubr()

```

Most have quite few missing sites. There are two with a negative number, both have very few snps (0 and 1) and thus some missing values anyway. But overall the number of sites makes sense when comparing it to end - start now.  

The number of SNPs correlates well with the number of sites. When we look at how the density might influence $F_{ST}$ it does not seem like it does when looking at nr of sites or region length. 

Reminder that $\pi$ and $d_{xy}$ are correlated:  

```{r pi dxy corr}
ggplot(Popstats, aes(x = pi_blue, y = dxy_blue_red))+
  geom_point(alpha = 0.6)+
  theme_pubr()
```


# Compare age categories

### Cold  

Let's first check the gene lengths to make sure that couldn't bias the results.

```{r cold age gene length, fig.show='hold',fig.height=3.5,fig.width=3}
PopstatsCold <- filter(Popstats, cvb != "not_sig")
test <- kruskal.test(sites ~ cvb, PopstatsCold)
ggplot(PopstatsCold, aes(x = cvb, y = sites))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter()+
  ggtitle(paste0("Kruskal p-value: ",round(test$p.value,4)))+
  theme_pubr(base_size = 10)
test <- kruskal.test(snps ~ cvb, PopstatsCold)
ggplot(PopstatsCold, aes(x = cvb, y = snps))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter()+
  ggtitle(paste0("Kruskal p-value: ",round(test$p.value,4)))+
  theme_pubr(base_size = 10)

```

There are no differences in number of sites or SNPs between categories.  

```{r cold age}
PopstatsCold <- filter(Popstats, cvb != "not_sig")
# are they significantly different?
test <- kruskal.test(Fst_blue_red ~ cvb, PopstatsCold) # p > 0.05 -> no not different
ggplot(PopstatsCold, aes(x = cvb, y = Fst_blue_red))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter()+
  ggtitle(paste0("Kruskal p-value: ",round(test$p.value,4)))+
  theme_pubr(base_size = 10)

# are they significantly different?
test <- kruskal.test(dxy_blue_red ~ cvb, PopstatsCold) # p > 0.05 -> no not different
ggplot(PopstatsCold, aes(x = cvb, y = dxy_blue_red))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter()+
  ggtitle(paste0("Kruskal p-value: ",round(test$p.value,4)))+
  theme_pubr(base_size = 10)

test <- kruskal.test(pi_red ~ cvb, PopstatsCold) # p > 0.05 -> no not different
ggplot(PopstatsCold, aes(x = cvb, y = pi_blue))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter()+
  ggtitle(paste0("Kruskal p-value: ",round(test$p.value,4)))+
  theme_pubr(base_size = 10)

```

### Hot  

Let's again first check the gene lengths to make sure that couldn't bias the results

```{r cold age gene length, fig.show='hold',fig.height=3.5,fig.width=3}
PopstatsHot <- filter(Popstats, hvb != "not_sig")
test <- kruskal.test(sites ~ cvb, PopstatsHot)
ggplot(PopstatsHot, aes(x = cvb, y = sites))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter()+
  ggtitle(paste0("Kruskal p-value: ",round(test$p.value,4)))+
  theme_pubr(base_size = 10)
test <- kruskal.test(snps ~ cvb, PopstatsHot)
ggplot(PopstatsHot, aes(x = cvb, y = snps))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter()+
  ggtitle(paste0("Kruskal p-value: ",round(test$p.value,4)))+
  theme_pubr(base_size = 10)

```

Huh there's a bit of a trend here with opposite direction genes being smaller/shorter.  

```{r hot age}
PopstatsHot <- filter(Popstats, hvb != "not_sig")
# are they significantly different?
test <- kruskal.test(Fst_blue_red ~ hvb, PopstatsHot) # p > 0.05 -> no not different
ggplot(PopstatsHot, aes(x = hvb, y = Fst_blue_red))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter()+
  ggtitle(paste0("Kruskal p-value: ",round(test$p.value,4)))+
  theme_pubr(base_size = 10)

# are they significantly different?
test <- kruskal.test(dxy_blue_red ~ hvb, PopstatsHot) # p > 0.05 -> no not different
ggplot(PopstatsHot, aes(x = hvb, y = dxy_blue_red))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter()+
  ggtitle(paste0("Kruskal p-value: ",round(test$p.value,4)))+
  theme_pubr(base_size = 10)
# there is a difference! But what?
dunnTest(dxy_blue_red ~ hvb, PopstatsHot, method = "bh") 

test <- kruskal.test(pi_blue ~ hvb, PopstatsHot) # p > 0.05 -> no not different
ggplot(PopstatsHot, aes(x = hvb, y = pi_blue))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter()+
  ggtitle(paste0("Kruskal p-value: ",round(test$p.value,4)))+
  theme_pubr(base_size = 10)
# there is a difference! But what?
dunnTest(pi_blue ~ hvb, PopstatsHot, method = "bh") 
PopstatsHot %>% group_by(hvb) %>% summarise(mean(pi_blue))

test <- kruskal.test(pi_red ~ hvb, PopstatsHot) # p > 0.05 -> no not different
ggplot(PopstatsHot, aes(x = hvb, y = pi_red))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter()+
  ggtitle(paste0("Kruskal p-value: ",round(test$p.value,4)))+
  theme_pubr(base_size = 10)
# there is a difference! But what?
dunnTest(pi_red ~ hvb, PopstatsHot, method = "bh") 

PopstatsHot %>% group_by(hvb) %>% summarise(mean(pi_red))
```

There are no significant differences between the cold genes, but in heat, there are differences in $d_{xy}$ and $\pi$. $D_{xy}$ is different in all comparisons except just_1week - opposite_direction, just_1week - same_direction and same_direction - same_slope. For $\pi_{red}$ the same comparisons are also significant, and also for $\pi_{blue}$ it's the same except just_1week - opposite_direction are also different.  

Opposite direction genes have the highest $\pi$ values, so the highest nucleotide diversity. Those genes are expressed differently in the two age groups in the heat, so the importance of their functions in relation to heat changes during development. We thus don't expect positive selection because selection pressure changes during development and thus likely cannot take effect. This absence of selection pressure is in line with the high nucleotide diversity we observe.  

Check whether they are same genes as hotup.

```{r check same genes}
goppdir_hot <- filter(PopstatsHot, hvb == "opposite_direction")$geneID
load("../ChamberTempExp/data/working/RNASeq_countdata/NewPHgenesForGO.RData")
goppdir_hot %in% hotup # no not at all
```

No there is no overlap.  

```{r plots for paper, include=FALSE}

# maybe filter out snps < 10
PopstatsHot <- filter(PopstatsHot, snps > 10)
test_dxy <- kruskal.test(dxy_blue_red ~ hvb, PopstatsHot) # p > 0.05 -> no not different
dunnTest(dxy_blue_red ~ hvb, PopstatsHot, method = "bh") 
ggplot(PopstatsHot, aes(x = hvb, y = dxy_blue_red, colour = hvb))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(size = 1.7, alpha = 0.7)+
  geom_signif(annotations = "*", y_position = 0.0135, xmin = 1, xmax = 1.95, tip_length = 0.01,textsize = 5, colour = "black")+ # just_1week - just_8week
  geom_signif(annotations = "*", y_position = 0.0135, xmin = 2.05, xmax = 2.95, tip_length = 0.01, textsize = 5, colour = "black")+ # just_8week - opposite_direction
  # geom_signif(annotations = "*", y_position = 0.014, xmin = 2.1, xmax = 4, tip_length = 0.01, textsize = 5, colour = "black")+ #  just_8week - same_direction
  geom_signif(annotations = "*", y_position = 0.0135, xmin = 3.05, xmax = 5, tip_length = 0.01, textsize = 5, colour = "black")+ #  opposite_direction - same_slope
  geom_signif(annotations = "*", y_position = 0.014, xmin = 1, xmax = 5, tip_length = 0.01, textsize = 5, colour = "black")+ # just_1week - same_slope
  # geom_signif(annotations = "NA", y_position = 0.0135, xmin = 1, xmax = 3, tip_length = 0.01, textsize = 3, colour = "black")+ # just_1week - opposite_direction
  # geom_signif(annotations = "NA", y_position = 0.014, xmin = 1, xmax = 3.95, tip_length = 0.01, textsize = 3, colour = "black")+ # just_1week - same_direction
  # geom_signif(annotations = "NA", y_position = 0.014, xmin = 4.05, xmax = 5, tip_length = 0.01, textsize = 3, colour = "black")+ # same_direction - same_slope
  scale_y_break(c(0.007, 0.013), space = 0.3)+
  geom_hline(yintercept = 0.013, linetype = 3, alpha = 0.3)+
  scale_y_continuous(limits = c(0, 0.0144))+
  #scale_colour_manual(values = c( "#e1bee7", "#ba68c8", "#9c27b0", "#6800a4", "#4a148c"))+
  scale_colour_manual(values = c( "#780852", "#780852", "#780852", "#780852", "#780852"))+
  scale_x_discrete(labels = c("Only 1-week-old", "Only 8-week-old","Opposite direction", "Same direction", "Not age modulated"))+
  guides(colour = guide_legend(nrow = 2))+
  xlab("Thermal response across age")+ylab("Absolute differentiation (Dxy)")+
  theme_pubr(base_size = 17, legend = "none")+
  theme(axis.text.y.right = element_blank(),
  axis.ticks.y.right = element_blank(),
  axis.line.y.right  = element_blank(),axis.text.x = element_text(angle = 20, hjust = 1))

## pi
# put them into one plot, so need to reshape
PopstatsHot_pilong <- pivot_longer(PopstatsHot, pi_black:pi_red, names_to = "pop", names_prefix = "pi_", values_to = "pi")

testpiB <- kruskal.test(pi_blue ~ hvb, PopstatsHot)
testpiR <- kruskal.test(pi_red ~ hvb, PopstatsHot) 
dunnTest(pi_blue ~ hvb, PopstatsHot, method = "bh") 
dunnTest(pi_red ~ hvb, PopstatsHot, method = "bh") 

ggplot(filter(PopstatsHot_pilong, pop != "black"), aes(x = hvb, y = pi, shape = pop, colour = hvb))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(size = 1.9, alpha = 0.7, position = position_jitterdodge())+
  geom_signif(annotations = "*", y_position = 0.0125, xmin = 0.8, xmax = 1.75, tip_length = 0.01, textsize = 5, colour = "blue")+ # just_1week - just_8week blue
  geom_signif(annotations = "*", y_position = 0.0125, xmin = 1.85, xmax = 2.75, tip_length = 0.01, textsize = 5, colour = "blue")+ # just_8week - opposite_direction blue
  geom_signif(annotations = "*", y_position = 0.013, xmin = 1.85, xmax = 3.75, tip_length = 0.01, textsize = 5, colour = "blue")+ # just_8week - same_direction blue
  geom_signif(annotations = "*", y_position = 0.0135, xmin = 0.8, xmax = 4.75, tip_length = 0.01, textsize = 5, colour = "blue")+ # just_1week - same_slope blue
  geom_signif(annotations = "*", y_position = 0.0125, xmin = 2.85, xmax = 4.75, tip_length = 0.01, textsize = 5, colour = "blue")+ # opposite_direction - same_slope blue
  geom_signif(annotations = "*", y_position = 0.0122, xmin = 3.2, xmax = 5.2, tip_length = 0.01, textsize = 5, colour = "red")+ # opposite_direction - same_slope red
  geom_hline(yintercept = 0.012, linetype = 3, alpha = 0.3)+
  scale_y_break(c(0.008, 0.012), space = 0.3)+
  scale_y_continuous(limits = c(0, 0.014))+
  #scale_colour_manual(values = c("#e1bee7", "#ba68c8", "#9c27b0", "#6800a4", "#4a148c"))+
  scale_colour_manual(values = c( "#780852", "#780852", "#780852", "#780852", "#780852"))+
  scale_x_discrete(labels = c("Only 1-week-old", "Only 8-week-old","Opposite direction", "Same direction", "Not age modulated"))+
  scale_shape_discrete(labels = c(expression(italic("S.c. australis")), expression(italic("S.c. massaicus"))))+
  guides(shape = guide_legend(position = "top", override.aes = list(linetype = c(0,0), size = c(2.5,2.5))),
         colour = "none")+
  xlab("Thermal response across age")+ylab("Nucleotide diversity (pi)")+
  theme_pubr(base_size = 17)+
  theme(axis.text.y.right = element_blank(),
  axis.ticks.y.right = element_blank(),
  axis.line.y.right  = element_blank(),axis.text.x = element_text(angle = 20, hjust = 1), legend.title = element_blank())

```


## Tajima's D

There is literature reporting balancing selection in heat shock proteins, mostly due to heterozygote advantage. Thus it could be worth it to corroborate this with some method. Tajima's D (the difference between the mean number of pairwise differences ($\pi$) and the number of segregating sites, each scaled so that they are expected to be the same in a neutrally evolving population of constant size) should be > 0 in the case of balancing selection, which means more haplotypes than nr of segregating sites, thus a lack of rare alleles.  

```{r get tajima d data, include=FALSE}
td_red <- read.table("data/FstTargeted/monobiallelic.500bp_start.ageint.red.Tajima.D", header = TRUE)
td_blue <- read.table("data/FstTargeted/monobiallelic.500bp_start.ageint.blue.Tajima.D", header = TRUE)
# delete nan rows
td_red <- filter(td_red, !is.na(TajimaD))
td_blue <- filter(td_blue, !is.na(TajimaD))

# for red
td_red$bin_end <- td_red$BIN_START + 350000
# add values to region overlapping genes in popstats, probably easiest to do scaffold by scaffold
matchmat <- sapply(Popstats$start, function(x) x > td_red$BIN_START & x < td_red$bin_end) # rows from td, columns from popstats
diag(matchmat) # in the beginning it matches up but then not any more except here and there
# so now I have to kind of match the row number to the column, or, for each column get the position of the TRUE(s)
# and it only counts if it's the same scaffold
which(matchmat[,1])
matches <- apply(matchmat,2, which)
nmatches <- rep(0,nrow(Popstats))
Popstats$td_red <- NA
Popstats$td_red_snps <- NA
for (i in 1:nrow(Popstats)){
  candidates <- matches[[i]]
  for (j in candidates){
    if (td_red[j,"CHROM"] == Popstats[i,"scaffold"]){
      Popstats[i,"td_red"] <- td_red[j,"TajimaD"]
      Popstats[i,"td_red_snps"] <- td_red[j,"N_SNPS"] # this is to test if this matches more or less
      nmatches[i] <- nmatches[i] + 1
    }
  }
}
nmatches


# check nr of snps
plot(Popstats$snps, Popstats$td_red_snps) 
Popstats$snp_diff <- Popstats$snps - Popstats$td_red_snps
plot(Popstats$sites,Popstats$snp_diff) 

## now for blue
td_blue$bin_end <- td_blue$BIN_START + 350000
# add values to region overlapping genes in popstats, probably easiest to do scaffold by scaffold
matchmat <- sapply(Popstats$start, function(x) x > td_blue$BIN_START & x < td_blue$bin_end) # rows from td, columns from popstats
diag(matchmat) #

which(matchmat[,1])
matches <- apply(matchmat,2, which)
nmatches <- rep(0,nrow(Popstats))
Popstats$td_blue <- NA
Popstats$td_blue_snps <- NA
for (i in 1:nrow(Popstats)){
  candidates <- matches[[i]]
  for (j in candidates){
    if (td_blue[j,"CHROM"] == Popstats[i,"scaffold"]){
      Popstats[i,"td_blue"] <- td_blue[j,"TajimaD"]
      Popstats[i,"td_blue_snps"] <- td_blue[j,"N_SNPS"] # this is to test if this matches more or less
      nmatches[i] <- nmatches[i] + 1
    }
  }
}
nmatches

## check nr of snps
plot(Popstats$snps, Popstats$td_blue_snps) # looks the same as before
Popstats$snp_diff <- Popstats$snps - Popstats$td_blue_snps
plot(Popstats$sites,Popstats$snp_diff) 
```

### In heat  

```{r plot tajimas d hot }
PopstatsHot <- filter(Popstats, hvb != "not_sig")
test <- kruskal.test(td_red ~ hvb, PopstatsHot) # p > 0.05 -> no not different
ggplot(PopstatsHot, aes(x = hvb, y = td_red))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter()+
  geom_abline(slope = 0, intercept = 0)+
  ggtitle(paste0("Kruskal p-value: ",round(test$p.value,4)))+
  theme_pubr(base_size = 10)
dunnTest(td_red ~ hvb, PopstatsHot, method = "bh") 

test <- kruskal.test(td_blue ~ hvb, PopstatsHot) # p > 0.05 -> no not different
ggplot(PopstatsHot, aes(x = hvb, y = td_blue))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter()+
  geom_abline(slope = 0, intercept = 0)+
  ggtitle(paste0("Kruskal p-value: ",round(test$p.value,4)))+
  theme_pubr(base_size = 10)
# here not different

```

In the blue subspecies, there does not seem to be any differences in Tajima's D in these categories. However, in the red population, Tajima's D is higher in genes that are only expressed in 8-week-olds (it's the only one different from 0). These genes had the overall *lowest* nucleotide diversity, so one would not expect balancing selection. Apparently it still has more haplotypes than segregating sites (lack of rare alleles). The oppositely expressed genes that we expected to have high Tajima's D have the second highest in the reds.  

The difference between reds and blues here is also apparent. Reds had generally lower nucleotide diversity than blues and are more inbred, which could maybe explain the generally higher Tajima's D/lack of rare alleles, as those could have gotten lost in the bottleneck of few founders on the farm. In general, one has to keep in mind when interpreting these results that they can be influenced by demography.  

Here is cold for completeness.

### In cold  

```{r plot tajimas d cold}
PopstatsCold <- filter(Popstats, cvb != "not_sig")
test <- kruskal.test(td_red ~ hvb, PopstatsCold) # p > 0.05 -> no not different
ggplot(PopstatsCold, aes(x = hvb, y = td_red))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter()+
  geom_abline(slope = 0, intercept = 0)+
  ggtitle(paste0("Kruskal p-value: ",round(test$p.value,4)))+
  theme_pubr(base_size = 10)
# probably to few to be significant

test <- kruskal.test(td_blue ~ hvb, PopstatsCold) # p > 0.05 -> no not different
ggplot(PopstatsCold, aes(x = hvb, y = td_blue))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter()+
  geom_abline(slope = 0, intercept = 0)+
  ggtitle(paste0("Kruskal p-value: ",round(test$p.value,4)))+
  theme_pubr(base_size = 10)
# here not different

```

The same trend emerges with cold genes, however, the differences are not significant.  

## Slightly different way of calculating Tajima's D

```{r get tajima d data 2, include=FALSE}
td_red <- read.table("data/FstTargeted/monobiallelic.500bp_start.ageint.red2.Tajima.D", header = TRUE)
td_blue <- read.table("data/FstTargeted/monobiallelic.500bp_start.ageint.blue2.Tajima.D", header = TRUE)
# delete nan rows
td_red <- filter(td_red, !is.na(TajimaD))
td_blue <- filter(td_blue, !is.na(TajimaD))

# for red
td_red$bin_end <- td_red$BIN_START + 10000
# add values to region overlapping genes in popstats
# now a bit more complicated because there could be more than one and not just start should be in it
# so I do start and end and then take all peaces in between later
matchmat_start <- sapply(Popstats$start, function(x) x > td_red$BIN_START & x < td_red$bin_end) # rows from td, columns from popstats
matchmat_end <- sapply(Popstats$end, function(x) x > td_red$BIN_START & x < td_red$bin_end) # rows from td, columns from popstats
# so now I have to kind of match the row number to the column, or, for each column get the position of the TRUE(s)
# and it only counts if it's the same scaffold
matches_start <- apply(matchmat_start,2, which)
matches_end <- apply(matchmat_end,2, which)
Popstats$td_red2 <- NA
Popstats$what <- NA
Popstats$start_cands <- 0
for (i in 1:nrow(Popstats)){
  candidates_start <- matches_start[[i]]
  candidates_end <- matches_end[[i]]
  for (j in candidates_start){
    if (td_red[j,"CHROM"] == Popstats[i,"scaffold"]){ # only for start on same scaffold (I expect there to be only one right candidate always)
      # find matching end position
      end <- candidates_end[which((j <= candidates_end & candidates_end < j + 40) | j == candidates_end)] # I expect end to be close, max 350000/10000 = 35 away, or the same
      if (length(end) > 1){
        if (td_red[end[1],"CHROM"] == Popstats[i,"scaffold"]){
          end <- end[1]
        } else if (td_red[end[2],"CHROM"] == Popstats[i,"scaffold"]){
          end <- end[2]
        }else{
          Popstats[i,"what"] <- "weird_end"
          next # I hope this makes that it will jump to the next i (not j) (but it never happened anyway)
        }
      } # if there are multpiple candidates, take which is on the same chromosome
      if (is.numeric(end) & length(end) > 0){
        Popstats[i,"td_red2"] <- mean(td_red[j:end,"TajimaD"]) # take the average of all rows in between (if they are the same it will still work)
        Popstats[i,"snps2"] <- sum(td_red[j:end,"N_SNPS"])
        Popstats[i,"what"] <- paste0("end_exists_",end)
      }else{ # else just take start
        Popstats[i,"td_red2"] <- td_red[j,"TajimaD"]
        Popstats[i,"snps2"] <- td_red[j,"N_SNPS"]
        Popstats[i,"what"] <- "end_missing"
      }
      # somehow record what happened
      Popstats[i,"start_cands"] <- Popstats[i,"start_cands"] +1
    }
  } # I guess it could also happen for small chromosome that only the end is in a chunk
  if (is.na(Popstats[i,"td_red2"])){ # so in that case (when start is missing) check for ends and take them
    for (e in candidates_end){
      if (td_red[e,"CHROM"] == Popstats[i,"scaffold"]){
        Popstats[i,"td_red2"] <- td_red[e,"TajimaD"]
        Popstats[i,"what"] <- "start_missing"
      }
    }
  }
}


## now for blue
td_blue$bin_end <- td_blue$BIN_START + 10000
# add values to region overlapping genes in popstats
# now a bit more complicated because there could be more than one and not just start should be in it
# so I do start and end and then take all peaces in between later
matchmat_start <- sapply(Popstats$start, function(x) x > td_blue$BIN_START & x < td_blue$bin_end) # rows from td, columns from popstats
matchmat_end <- sapply(Popstats$end, function(x) x > td_blue$BIN_START & x < td_blue$bin_end) # rows from td, columns from popstats
# so now I have to kind of match the row number to the column, or, for each column get the position of the TRUE(s)
# and it only counts if it's the same scaffold
matches_start <- apply(matchmat_start,2, which)
matches_end <- apply(matchmat_end,2, which)
Popstats$td_blue2 <- NA
Popstats$what <- NA
Popstats$start_cands <- 0
for (i in 1:nrow(Popstats)){
  candidates_start <- matches_start[[i]]
  candidates_end <- matches_end[[i]]
  for (j in candidates_start){
    if (td_blue[j,"CHROM"] == Popstats[i,"scaffold"]){ # only for start on same scaffold (I expect there to be only one right candidate always)
      # find matching end position
      end <- candidates_end[which((j <= candidates_end & candidates_end < j + 40) | j == candidates_end)] # I expect end to be close, max 350000/10000 = 35 away, or the same
      if (length(end) > 1){
        if (td_blue[end[1],"CHROM"] == Popstats[i,"scaffold"]){
          end <- end[1]
        } else if (td_blue[end[2],"CHROM"] == Popstats[i,"scaffold"]){
          end <- end[2]
        }else{
          Popstats[i,"what"] <- "weird_end"
          next # I hope this makes that it will jump to the next i (not j) (but it never happened anyway)
        }
      } # if there are multpiple candidates, take which is on the same chromosome
      if (is.numeric(end) & length(end) > 0){
        Popstats[i,"td_blue2"] <- mean(td_blue[j:end,"TajimaD"]) # take the average of all rows in between (if they are the same it will still work)
        Popstats[i,"what"] <- paste0("end_exists_",end)
      }else{ # else just take start
        Popstats[i,"td_blue2"] <- td_blue[j,"TajimaD"]
        Popstats[i,"what"] <- "end_missing"
      }
      # somehow record what happened
      Popstats[i,"start_cands"] <- Popstats[i,"start_cands"] +1
    }
  } #It could also happen for small chromosome that only the end is in a chunk
  if (is.na(Popstats[i,"td_blue2"])){ # so in that case (when start is missing) check for ends and take them
    for (e in candidates_end){
      if (td_blue[e,"CHROM"] == Popstats[i,"scaffold"]){
        Popstats[i,"td_blue2"] <- td_blue[e,"TajimaD"]
        Popstats[i,"what"] <- "start_missing"
      }
    }
  }
}

```

```{r compare Tajimas D}
plot(Popstats$td_blue,Popstats$td_blue2)
abline(0,1)
plot(Popstats$td_red,Popstats$td_red2)
abline(0,1)
```

This other method seems overall similar and there doesn't seem to be a bias. However, some of the differences are quite big. Let's see if this influences the results.  

### TD in heat new  

```{r plot tajimas d hot new }
PopstatsHot <- filter(Popstats, hvb != "not_sig")
test <- kruskal.test(td_red2 ~ hvb, PopstatsHot) # p > 0.05 -> no not different
ggplot(PopstatsHot, aes(x = hvb, y = td_red2))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter()+
  geom_abline(slope = 0, intercept = 0)+
  ggtitle(paste0("Kruskal p-value: ",round(test$p.value,4)))+
  theme_pubr(base_size = 10)
# there is no difference any more

# what with filtering out ones with few snps
PopstatsHot <- filter(PopstatsHot, snps2 > 10)
test <- kruskal.test(td_red2 ~ hvb, PopstatsHot) # p > 0.05 -> no not different
ggplot(PopstatsHot, aes(x = hvb, y = td_red2))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter()+
  geom_abline(slope = 0, intercept = 0)+
  ggtitle(paste0("Kruskal p-value: ",round(test$p.value,4)))+
  theme_pubr(base_size = 10)
# makes it less significant

test <- kruskal.test(td_blue2 ~ hvb, PopstatsHot) # p > 0.05 -> no not different
ggplot(PopstatsHot, aes(x = hvb, y = td_blue2))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter()+
  geom_abline(slope = 0, intercept = 0)+
  ggtitle(paste0("Kruskal p-value: ",round(test$p.value,4)))+
  theme_pubr(base_size = 10)
# here not different

```

The trends are similar to before, however the result is not significant now. In cold there also no big change.  

```{r plot tajimas d 2 cold, include=FALSE}
PopstatsCold <- filter(Popstats, cvb != "not_sig")
test <- kruskal.test(td_red2 ~ hvb, PopstatsCold) # p > 0.05 -> no not different
ggplot(PopstatsCold, aes(x = hvb, y = td_red2))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter()+
  geom_abline(slope = 0, intercept = 0)+
  ggtitle(paste0("Kruskal p-value: ",round(test$p.value,4)))+
  theme_pubr(base_size = 10)
# probably to few to be significant

test <- kruskal.test(td_blue2 ~ hvb, PopstatsCold) # p > 0.05 -> no not different
ggplot(PopstatsCold, aes(x = hvb, y = td_blue2))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter()+
  geom_abline(slope = 0, intercept = 0)+
  ggtitle(paste0("Kruskal p-value: ",round(test$p.value,4)))+
  theme_pubr(base_size = 10)
# here not different

```

```{r save popstats}
save(Popstats, file = "data/FstTargeted/Popstats_ageint.RData")
```


## Hot up and down ages

There were especially many genes up and down regulated in heat only in 1-week-olds, so maybe something special is happening when they are very young. Let's have a look at those genes.  

```{r get age up down data,fig.show='hold',fig.height=3.5,fig.width=3}
# need to get that info from somewhere first
load("../ChamberTempExp/data/working/RNASeq_countdata/cat_eff_age_resp.RData")
# add up down info to Popstats
Popstats <- left_join(Popstats, dplyr::select(cat_eff2, c(gene, week1, week8)), by = join_by(geneID == gene)) # doubles some genes with different effects in ages
# simplify to just up down
Popstats$week1_dir <- ifelse(Popstats$week1 < 0, "down","up")
Popstats$week8_dir <- ifelse(Popstats$week8 < 0, "down","up")
# make up vs down plot by age, just hot
Popstats_hotw1 <-  filter(Popstats, hvb == "just_1week" & !is.na(week1_dir))
library(brunnermunzel)
test <- brunnermunzel.test(filter(Popstats_hotw1, week1_dir == "up")$dxy_blue_red, filter(Popstats_hotw1, week1_dir == "down")$dxy_blue_red)
ggplot(Popstats_hotw1, aes(x = week1_dir, y = dxy_blue_red))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(alpha = 0.6)+
  ggtitle(paste("Brunner Munzel p-value:",round(test$p.value,6)))+
  theme_pubr(base_size = 10)

Popstats_hotw8 <-  filter(Popstats, hvb == "just_8week" & !is.na(week8_dir))
library(brunnermunzel)
test <- brunnermunzel.test(filter(Popstats_hotw8, week8_dir == "up")$dxy_blue_red, filter(Popstats_hotw8, week8_dir == "down")$dxy_blue_red)
ggplot(Popstats_hotw8, aes(x = week8_dir, y = dxy_blue_red))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(alpha = 0.6)+
  ggtitle(paste("Brunner Munzel p-value:",round(test$p.value,6)))+
  theme_pubr(base_size = 10)

```

We cannot conclude that differences in age drive the difference in $d_{xy}$ in hot up to the rest/hot down. If our hypothesis that the higher $\pi$ in hot up is due to heat shock proteins that are apparently known to be under balancing selection is true, this makes sense, because HSPs are expressed in both ages, not just 1-week-olds.  

# Check gene length thing

### Fig 17A: number of sites per gene in genes with age:treatment interaction

```{r oppdir noint pvals}
ggplot(PopstatsHot, aes(x = hvb, y = sites, colour = hvb))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(size = 1.7, alpha = 0.7)+
  scale_colour_manual(values = c( "#ff0000", "#f8c75d", "#f98400", "#00a08a", "#5bbcd6" ),
                      labels = c("just 1-week-old", "just 8-week-old","opposite direction", "same direction", "no age interaction"))+
  scale_x_discrete(labels = c("just 1-week-old", "just 8-week-old","opposite direction", "same direction", "no age interaction"))+
  guides(colour = guide_legend(nrow = 2))+
  # labs(subtitle= paste0("Kruskal p-value: ",round(test_dxy$p.value,4)))+
  xlab("")+ylab("Nr of sites in gene")+
  theme_pubr()+
  theme(axis.text.x = element_text(angle = 20, hjust = 1), legend.title = element_blank())
```

### Fig 17B: Brunnermunzel p-values of pi and dxy between just_8week and just_1week after sampling a similar distribution of number of sites from just_1week as just_8week

```{r just1w just8w pvals}
stats_1w8w <- get_similar_dist(Popstats_int, "hvb", "just_8week","just_1week",20)[[1]]
pvals_just1w_just8w <- get_similar_dist(Popstats_int, "hvb", "just_8week","just_1week",20)[[2]]
pi_red <- length(which(pvals_just1w_just8w$pi_red < 0.05))
pi_blue <- length(which(pvals_just1w_just8w$pi_blue < 0.05))
dxy <- length(which(pvals_just1w_just8w$dxy < 0.05))

ggplot(stats_1w8w, aes(x = cat4, y = dxy_blue_red))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(alpha = 0.7)+
  annotate("text", x=2, y = 0.003, label = paste0("P values below 0.05 for Pi red: ", pi_red,"/20"))+
  annotate("text", x=2, y = 0.0035, label = paste0("P values below 0.05 for Pi blue: ", pi_blue,"/20"))+
  annotate("text", x=2, y = 0.0025, label = paste0("P values below 0.05 for dxy: ", dxy,"/20"))+
  theme_pubr(base_size = 10)

```

### Fig 17C: Brunnermunzel p-values of pi and dxy between "no age interaction" and "opposite direction" after sampling a similar distribution of number of sites from "no age interaction" as "opposite direction"

```{r opp dir same slope pvals}
stats_oppsame <- get_similar_dist(Popstats_int, "hvb", "opposite_direction","same_slope",20)[[1]]
pvals_oppdir_noint <- get_similar_dist(Popstats_int, "hvb", "opposite_direction","same_slope",20)[[2]]
pi_red <- length(which(pvals_oppdir_noint$pi_red < 0.05))
pi_blue <- length(which(pvals_oppdir_noint$pi_blue < 0.05))
dxy <- length(which(pvals_oppdir_noint$dxy < 0.05))

ggplot(stats_oppsame, aes(x = cat4, y = dxy_blue_red))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(alpha = 0.7)+
  annotate("text", x=2, y = 0.006, label = paste0("P values below 0.05 for Pi red: ", pi_red,"/20"))+
  annotate("text", x=2, y = 0.005, label = paste0("P values below 0.05 for Pi blue: ", pi_blue,"/20"))+
  annotate("text", x=2, y = 0.004, label = paste0("P values below 0.05 for dxy: ", dxy,"/20"))+
  theme_pubr(base_size = 10)

```



